# Washington State Domestic Violence Fatalities

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width = 9, fig.height = 7, fig.path = "../figs/",
                      warning = FALSE, message = FALSE, echo = FALSE)
```

I am interested in understanding trends in domestic violence fatalities in
Washington state, specifically:

* Have rates of domestic violence homicides, suicides, and fatal police
  interventions changed over time?
* Which counties have the highest numbers of domestic violence homicides and
  suicides (particularly after normalizing for population)?


To address these questions I am using data from the Washington State Coalition
Against Domestic Violence's
[Domestic Violence Fatality Review](http://dvfatalityreview.org/) and population
data from the Office of Financial Management. The Domestic Violence Fatality
Review provides county-level data on domestic violence homicides, suicides, and
fatal police interventions from 1997 through 2015 (though I will only be using
data through 2014 as the second half of 2015 is not yet available). I used
[Tabula](http://tabula.technology/) to convert the
[data PDF](http://dvfatalityreview.org/2015/12/30/updated-through-6302015-washington-state-domestic-violence-fatalities-by-county/)
into a CSV. The resulting CSV is still quite messy and required some significant
data cleaning. The Office of Financial Management publishes yearly county-level
population estimates which are probably good enough for these purposes.

All of the cleaning and plotting of the data (with the exception of running the
data PDF through Tabula to get a CSV) was done in R. To reduce clutter and focus
on the visualizations I've hidden the code chunks, but they can be viewed in
[the .Rmd file](https://github.com/karawoo/INFX-598J-assignment-2/blob/master/code/dv_vis.Rmd).

```{r dv_data}
## First load the required packages
library("readxl")
library("tidyr")
library("dplyr")
library("maps")
library("ggplot2")
library("viridis")

## Load data without the header row for now -- we'll add that back in later
dv <- read.table(
  "../data/original/tabula-fatalities-by-county-through-6-30-2015.csv",
  sep = ",",
  skip = 1,
  header = FALSE,
  stringsAsFactors = FALSE
)

## Fix row of fatality types (homicide, suicide, police intervention)
dv[1, ] <- gsub("\n", "", dv[1, ])    # remove \n's
dv[1, grep("(?=.*h)(?=.*o)(?=.*m)(?=.*i)(?=.*c)(?=.*d)(?=.*e)",
            dv[1, ], perl = TRUE)] <- "homicide"
dv[1, grep("(?=.*s)(?=.*u)(?=.*i)(?=.*c)(?=.*d)(?=.*e)",
            dv[1, ], perl = TRUE)] <- "suicide"
dv[1, grep("(?=.*p)(?=.*o)(?=.*l)(?=.*i)(?=.*c)(?=.*e)",
            dv[1, ], perl = TRUE)] <- "police.intervention"

## Get years from first row of CSV
years <- scan(
  "../data/original/tabula-fatalities-by-county-through-6-30-2015.csv",
  nlines = 1, sep = ",", what = character()
)

## Fill in empty strings. There should be 3 entries for each year. This creates
## 6 for 2015; the last three are actually columns of totals that are going to
## be removed anyway
for (i in seq_along(years)) {
  if (i == 1) {
    years[i] <- years[i]
  }
  else if (years[i] == "") {
    years[i] <- years[i - 1]
  }
}

## Paste together year and fatality type into one column header -- these will be
## separated later after reshaping
newnames <- paste(years, dv[1, ], sep = "_")
newnames[1] <- "county"

## Use year and fatality type as column names
names(dv) <- newnames

## Now get rid of first row of data that contains fatality types, the last row
## of data that contains state-wide totals, and last three columns which are
## totals for the time series
dv <- dv[-c(1, nrow(dv)), -c(ncol(dv), ncol(dv) - 1, ncol(dv) - 2)]
```

```{r dv_reshape}
dv_long <- dv %>%
  ## Gather into long format
  gather(key = yeartype, value = fatalities, -county) %>%
  ## Separate yeartype column into year and fatality type
  separate(yeartype, c("year", "type"), sep = "_") %>%
  ## Based on the original PDF it looks like NAs should be true zeroes
  mutate(fatalities = ifelse(fatalities == "", "0", fatalities)) %>%
  ## Convert fatalities to numeric
  mutate(fatalities = as.numeric(fatalities))
```

```{r pop_data}
## Download OFM data
## download.file(
##   url = "http://www.ofm.wa.gov/pop/april1/hseries/ofm_april1_postcensal_estimates_pop_1960-present.xlsx",
##   destfile = "../data/original/postcensal_pop_estimates.xlsx"
## )

## Load population estimates
pop <- read_excel(
  "../data/original/postcensal_pop_estimates.xlsx",
  sheet = 2, skip = 3
)

## Make syntactically valid column names
names(pop) <- make.names(names(pop), unique = TRUE)

## Remove empty and unnecessary columns and rows
pop <- pop %>%
  select(-starts_with("NA")) %>%
  filter(Filter == 1 & !County %in% c("State", ".")) %>%
  select(-Line, -Filter, -Jurisdiction)
    
## Remove periods from column names
names(pop) <- gsub("\\.", "", names(pop))

## Convert data to long format
pop_long <- pop %>%
  gather(key = year, value = pop, -County) %>%
  rename(county = County) %>%
  ## Remove postcensal estimates in years where there is an actual census
  filter(!year %in% c("X1960PostcensalEstimateofTotalPopulation",
                      "X1970PostcensalEstimateofTotalPopulation",
                      "X1980PostcensalEstimateofTotalPopulation",
                      "X1990PostcensalEstimateofTotalPopulation",
                      "X2000PostcensalEstimateofTotalPopulation",
                      "X2010PostcensalEstimateofTotalPopulation")) %>%
  mutate(year = substr(year, 2, 5)) %>%
  ## Remove data from before 1997 (the start of DV data)
  filter(year >= 1997) %>%
  ## Make population and year numeric
  mutate(pop = as.numeric(pop),
         year = as.numeric(year))
```

```{r merge_data}
dat <- merge(dv_long, pop_long, by = c("county", "year"))
```

\newpage

## Distributions

What is the distribution of fatalities by county in a given year, normalized by
population?

```{r fatality_hist_by_pop, fig.width = 6, fig.height = 6}
dat %>%
  group_by(county, year) %>%
  summarize(tot_fatal_pop = sum(fatalities, na.rm = TRUE) / (unique(pop) / 10000)) %>%
  ggplot(aes(x = tot_fatal_pop)) +
  geom_histogram(binwidth = 0.1) +
  facet_wrap(~ year, ncol = 3) +
  ylab("Count") +
  xlab("Number of fatalities per 10,000 population") +
  ggtitle("Domestic Violence Fatalities")
```

\newpage

## Time series and temporal trends

What, if any, temporal trends are there in total fatalities and the three types
of fatalities?

```{r fatalities_line}
dat_total <- dat %>%
  group_by(year) %>%
  summarize(tot_fatal_state = sum(fatalities, na.rm = TRUE))

ggplot(dat_total, aes(x = year, y = tot_fatal_state, group = 1)) +
  geom_line() +
  ylab("Fatalities") +
  xlab("Year") +
  ggtitle("Total Domestic Violence Fatalities in Washington State")
```

```{r fatalities_line_by_type}
dat %>%
  group_by(year, type) %>%
  summarize(tot_fatal_state = sum(fatalities, na.rm = TRUE)) %>%
    ggplot(aes(x = year, y = tot_fatal_state, group = 1)) +
  geom_line() +
  facet_wrap(~ type, ncol = 1) +
  ylab("Fatalities") +
  xlab("Year") +
  ggtitle("Domestic Violence Fatalities in Washington State by Type")
```

\newpage 

If I were planning more confirmatory statistics with this data I'd probably want
to know about the degree of temporal autocorrelation.

```{r autocorrelation_total}
acf(dat_total$tot_fatal_state, main = "Autocorrelation function for statewide \ntotal domestic violence fatalities")
```

\newpage

Can we see any trends in homicides by county?

```{r homicide_line_color}
## Line graph of homicides
dat %>%
  filter(type == "homicide") %>%
  ggplot(aes(x = year, y = fatalities, group = county, color = county)) +
  geom_line()
```

Okay not that way...how about faceting?

```{r homicide_line_facet}
## Faceted by county
dat %>%
  filter(type == "homicide") %>%
  ggplot(aes(x = year, y = fatalities)) +
  facet_wrap(~ county, ncol = 5) +
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of fatalities") +
  xlab("Year") +
  ggtitle("Domestic Violence Homicide Fatalities")

```

\newpage

## Maps of fatalities by county

```{r homicide_map_data} 
## To visualize homicides on a map, we first need to acquire county boundary
## data from `ggplot2`'s `map_data()` function and merge it with our homicide
## data.

## Gather county spatial data
county_map <- ggplot2::map_data("county") %>%
  filter(region == "washington") %>%
  ## Rename subregion column to county
  rename(county = subregion) %>%
  ## Remove region column (washington)
  select(-region) %>%
  ## Capitalize counties
  mutate(county = gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2",
                     county, perl = TRUE))

## Merge homicide data with county map
hom_county <- dat %>%
  filter(type == "homicide") %>%
  merge(county_map, by = "county")

```

```{r homicide_map_total, fig.height = 8.5}
## Map of number of homicides in each county by year
ggplot(hom_county, aes(x = long, y = lat, group = group)) +
  facet_wrap(~ year, ncol = 3) +
  geom_polygon(colour = "grey", size = 0.3, aes(fill = fatalities)) +
  scale_fill_viridis(alpha = 1, begin = 0, end = 1, discrete = FALSE,
                     option = "D") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  expand_limits(x = hom_county$long, y = hom_county$lat) +
  coord_map("polyconic") +
  labs(fill="Number of \nhomicides") +
  ggtitle("Domestic Violence Homicides")
```

Predictably, this basically looks like a population map of Washington. So
instead let's look at the number of homicides per 10,000 people.

```{r homicide_map_by_pop, fig.height = 8.5}
## Map of number of homicides/population by year
ggplot(hom_county, aes(x = long, y = lat, group = group)) +
  facet_wrap(~ year, ncol = 3) +
  geom_polygon(colour = "grey", size = 0.3, aes(fill = fatalities / (pop / 10000))) +
  scale_fill_viridis(alpha = 1, begin = 0, end = 1, discrete = FALSE,
                     option = "D") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  expand_limits(x = hom_county$long, y = hom_county$lat) +
  coord_map("polyconic") +
  labs(fill="Homicides per \n10,000 population") +
  ggtitle("Domestic Violence Homicides")
```
